A weakly structured stem for human origins in Africa

### 2022-12-23 paper.tex
Updated Discussion

We quantify migration rates among populations after their divergence
$\approx120ka$. Prior to the Bantu expansion, migration between the ancestors
of the Nama and other groups is an order of magnitude weaker than that observed
between between western and eastern Africans (Table \ref{tab:migration-rates}).
All models infer relatively high gene flow between Ethiopia and western Africa
($m\approx2\times10^{-4}$, the constant proportion of immigrant lineages per
generation over 60ka). We further find that Back-to-Africa gene flow at the
beginning of the Holocene primarily affected the ancestors of the Ethiopian
agricultural populations \citep{Molinaro2019-rf}, comprising over half of their
genetic ancestry, estimated to be 64--65\%.  We observe significant gene flow
from the Amhara and Oromo into the Nama, a signal which is likely a proxy for
the movement of eastern African caprid and cattle pastoralists
\citep{Henn2008-xo,Breton2014-xb}, here estimated to constitute a 25\% ancestry
contribution 2,000 ya. While this gene flow is not apparent from the ADMIXTURE
plot (Figure~\ref{fig:diversity}), the ancestry is likely grouped into the
Khoe-San component which has drifted appreciably from its ancestral eastern
African source.  Colonial period admixture from Europeans into the Nama was
estimated at 15\%, similar to proportions suggested by \texttt{ADMIXTURE}
(Figure~\ref{fig:diversity}).

@@

Any attempt at building detailed models of human history is subject to model
misspecification. This is true of earlier studies, which often assumed that
data inconsistent with a single-origin model must be explained by archaic
hominin admixture. This is also true of this study. While it remains
prohibitive to fully explore the space of plausible models of early human
population structure, we sought to capture model uncertainty by exploring
multiple parameterizations of early history. The best-fit models presented
here include reticulation and migration between early human populations rather
than archaic hominin admixture from long-isolated branches. We cannot rule out
that more complex models involving additional stems, continuous structure, or
hybrid models including both weak structure and archaic hominin admixture may
better explain the data. Because parameters related to the split time,
migration rates, and relative sizes of the early stems were variable across
models, reflecting a degree of confounding among these parameters, we refrained
from introducing additional branches associated with more parameters during
that period.  Rather than interpreting the two stems as representing
well-defined and stable populations over hundreds of thousands of years, we
interpret the weakly structured stem as consistent with a population
coalescence and fragmentation model \citep{Scerri2019-xg}. Models including
additional diversity within Africa, and early ancient DNA samples from Africa,
could further distinguish the archaic hominin admixture model from the
weakly-structured-stem model.

@@

By contrast, our inferred models paint a more consistent picture of the Middle to Late Pleistocene as a critical period of change, assuming that estimates from
the recombination clock accurately relate to geological chronologies
(SI Section~\ref{sec:supp-mutation}).
During the late Middle Pleistocene, the multiple merger model indicates three
major stem lineages in Africa, tentatively assigned to southern (Stem 1S),
eastern (Stem 1E) and western/central Africa (Stem 2). While the length of
isolation among the stems is variable across model fits, models with a period of 
divergence, isolation and then a merger event (i.e. a ``reticulation'')
out-performed models with bifurcating divergence and continuous gene flow.

>>>>

We quantify migration rates among populations after their divergence
$\approx120ka$. Prior to agripastoralist expansion 5ka, migration between the ancestors
of the Nama and other groups is an order of magnitude weaker than that observed
between between western and eastern Africans (Table \ref{tab:migration-rates}).
All models infer relatively high gene flow between Ethiopia and western Africa
($m\approx2\times10^{-4}$, the constant proportion of immigrant lineages per
generation over 60ka). We further find that Back-to-Africa gene flow at the
beginning of the Holocene primarily affected the ancestors of the Ethiopian
agricultural populations \citep{Molinaro2019-rf}, comprising over half of their
genetic ancestry, estimated to be 64--65\%.  We observe significant gene flow
from the Amhara and Oromo into the Nama, a signal which is likely a proxy for
the movement of eastern African caprid and cattle pastoralists
\citep{Henn2008-xo,Breton2014-xb}, here estimated to constitute a 25\% ancestry
contribution 2,000 ya. While this gene flow is not apparent from the ADMIXTURE
plot (Figure~\ref{fig:diversity}), the ancestry is likely grouped into the
Khoe-San component which has drifted appreciably from its ancestral eastern
African source.  Colonial period admixture from Europeans into the Nama was
estimated at 15\%, similar to proportions suggested by \texttt{ADMIXTURE}
(Figure~\ref{fig:diversity}).

@@

Any attempt at building detailed models of human history is subject to model
misspecification. This is true of earlier studies, which often assumed that
data inconsistent with a single-origin model must be explained by archaic
hominin admixture. This is also true of this study. While it remains
prohibitive to fully explore the space of plausible models of early human
population structure, we sought to capture model uncertainty by exploring
multiple parameterizations of early history. The best fit models presented
here include reticulation and migration between early human populations rather
than archaic hominin admixture from long-isolated branches (Figure~\ref{fig:proposed-models}C). 
Elements of a recent single-origin or African multi-regionalism (Figure~\ref{fig:proposed-models}A,D)
feature in our best fit models, as indicated in the recent time of contemporary population divergence
or the gene flow between disparate stems, respectively.
We caution that we cannot rule out that more complex models involving additional stems, continuous structure, or
hybrid models including both weak structure and archaic hominin admixture may
better explain the data. Because parameters related to the split time,
migration rates, and relative sizes of the early stems were variable across
models, reflecting a degree of confounding among these parameters, we refrained
from introducing additional branches associated with more parameters during
that period.  Rather than interpreting the two stems as representing
well-defined and stable populations over hundreds of thousands of years, we
interpret the weakly structured stem as consistent with a population
coalescence and fragmentation model \citep{Scerri2019-xg}.

Additional African populations such as those 
from Central Africa, or pre-Holocene ancient DNA samples, could further test our .

@@

Our inferred models paint a more consistent picture of the Middle to Late Pleistocene as a 
critical period of change, assuming that estimates from the recombination clock accurately relate to geological chronologies
(SI Section~\ref{sec:supp-mutation}). 
During the late Middle Pleistocene, the multiple merger model indicates three
major stem lineages in Africa, tentatively assigned to southern (Stem 1S),
eastern (Stem 1E) and western/central Africa (Stem 2). While the length of
isolation among the stems is variable across model fits, models with a period of 
divergence, isolation and then a merger event (i.e. a ``reticulation'')
out-performed models with bifurcating divergence and continuous gene flow.
